### 2013-07-22 Tex/chapters/methods.tex
Bildunterschriften in einheitlichem case
Ergänzung zu themodyn. Theorie

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    To examine the disordered Ising model in Equilibrium,
    Its temperature \(T\),
    volume and number of particles \(N\) are constant, so that it is a canonical
    ensemble. In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \[p_i \propto e^{-\frac{\hat{H}}{k_B T}}\]
    Further the free energy \(F\) of an canonical
    ensemble is minimized and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g. \cite{nolting2005}).
    To determine \(F\), one has to know the partition function
    \[Z=\sum_i e^{-\frac{\hat{H}_i}{kT}}\]
    where the sum goes over all possible states \(i\) of the system.
    For small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But then there are of course next to
    infinity many different states, such that it is unfeasible to
    calculate the energy for each state, except for cases, where it is
    possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations.\\

@@

    \subsubsection{Equilibration- and Autocorrelation Time}
    \label{sssec:eqtime}
        To generate an equilibrium state one starts with an arbitrary state
        and waits until it is equilibrated. The count of sweeps till
        equilibrium is called \emph{equilibration time} \(t_{eq}\).
        All measurement should start after this time.\\
        To determine when the system is in equilibrium, one can watch the
        development of some observables and take the point at which there
        are no big changes anymore as equilibrium. In fig.
        \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
        the equilibrium is reached after approximately \(N=50\) sweeps for
        either an initial condition of all spins up and all spins random. It
        does not harm to double that value to be save.
        \begin{figure}[htbp]
            \centering
            \subfigure[Example of an equilibrating Ising system][]{
                    \label{sfig:equiandauto:equiE}
                    \includegraphics[width=0.47\textwidth]{plots/equiE}
            }
            \subfigure[Example of the autocorrelation of an Ising system][]{
                    \label{sfig:equiandauto:autoM}
                    \includegraphics[width=0.47\textwidth]{plots/autoM}
            }
            \caption[Examples for equilibration and autocorrelation]
            {
                \subref{sfig:equiandauto:equiE} Example of a Ising system
                    \(L=64\) reaching equilibrium at \(T=2.36\) and
                \subref{sfig:equiandauto:autoM} the autocorrelation of an
                    Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                    sweeps -- otherwise the decline is too steep to show)
                    on half logarithmic axis.
                    The straight line is an exponential fit \(\exp(-t/\tau)\)
                    with \(\tau = 342(1)\).
            }
            \label{fig:equiandauto}
        \end{figure}\\
        Because every state is generated from the state before, measurements
        of subsequent states are correlated. To determine when two states
        are independent, one calculates the normalized autocorrelation function
        \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
        \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
        logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
        To get the autocorrelation time one can either fit a exponential
        function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
        or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
        \(\tau\) is an estimate after which time two samples are not
        correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
        To make sure that the error is not underestimated one should wait
        \(2\tau\) sweeps between two measurements.
        The autocorrelation time is of course dependent on the temperature.
        For example for the standard Metropolis algorithm the fluctuations
        are strong at high temperatures and subsequent
        states are more dissimilar and therefore less correlated than at low
        temperatures, where less spins are flipping. But the longest
        autocorrelation times are encountered at the critical temperature.
        This effect is called \emph{critical slowing down} and is
        characterized by the \emph{dynamical critical exponent} \(z\)
        \cite{SwendsenWang1987}. The dependence of the autocorrelation time
        \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).
        As mentioned in section \ref{sssec:wolff}, the Wolff cluster algorithm
        decreases \(z\) dramatically. This causes the course of the autocorrelation
        time in dependence on temperature to change significantly as shown in
        section \ref{sec:results}.

>>>>

\subsection{Thermodynamic Theory}
\label{ssec:theory}
    %~ To examine the disordered Ising model in Equilibrium,
    %~ Its temperature \(T\),
    %~ volume and number of particles \(N\) are constant, so that it is a canonical
    %~ ensemble.
    In a canonical ensemble the probability \(p_i\) of a state
    \(i\) is distributed according to a Boltzman distribution
    \[p_i \propto e^{-\frac{\hat{H}}{k_B T}}\]
    Further the free energy \(F\) of an canonical
    ensemble is minimized and all observables can be derived from \(F\)
    in a straight forward way, as stated in every textbook about
    statistical physics (e.g. \cite{nolting2005}).\\
    Because of \(F=U-TS\) where \(U=\avg{H}\) is the internal energy and
    \(S\) entropy, one can guess, that for low \(T\) the internal energy
    will be low, and for big \(T\) the entropy will be high, to minimize
    \(F\). The Hamiltonian of the Ising model creates an conflict between
    \(U\) and \(S\) by assigning low energies to states with high order.
    So \(U\) is low if all spins are aligned, where \(S\) as a measure of
    disorder is low. This preliminary considerations make a phase
    transition at some \(T\), where a high entropy lowers \(F\) more than
    a low internal energy, very plausible.\\\
    To determine
    \[F=-k_{B}T \ln{Z}\]
    one has to know the partition function
    \[Z=\sum_i e^{-\frac{\hat{H}_i}{k_{B}T}}\]
    where the sum goes over all possible states \(i\) of the system.
    For small \(N\) the partition function is computable, but the system
    may show very different properties than in the thermodynamic limit.
    To minimize these finite size effects, it is desirable to examine
    systems with large \(N\). But then there are of course next to
    infinity many different states, such that it is unfeasible to
    calculate the energy for each state, except for cases, where it is
    possible to solve it analytically.\\
    If not, one can get estimates of the observables for big \(N\) using
    Monte Carlo simulations.\\

@@

    \subsubsection{Equilibration- and Autocorrelation Time}
    \label{sssec:eqtime}
        To generate an equilibrium state one starts with an arbitrary state
        and waits until it is equilibrated. The count of sweeps till
        equilibrium is called \emph{equilibration time} \(t_{eq}\).
        All measurement should start after this time.\\
        To determine when the system is in equilibrium, one can watch the
        development of some observables and take the point at which there
        are no big changes anymore as equilibrium. In fig.
        \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
        the equilibrium is reached after approximately \(N=50\) sweeps for
        either an initial condition of all spins up and all spins random. It
        does not harm to double that value to be save.
        \begin{figure}[htbp]
            \centering
            \subfigure[Example of an Equilibrating Ising System][]{
                    \label{sfig:equiandauto:equiE}
                    \includegraphics[width=0.47\textwidth]{plots/equiE}
            }
            \subfigure[Example of the Autocorrelation of an Ising System][]{
                    \label{sfig:equiandauto:autoM}
                    \includegraphics[width=0.47\textwidth]{plots/autoM}
            }
            \caption[Examples for Equilibration and Autocorrelation]
            {
                \subref{sfig:equiandauto:equiE} Example of a Ising system
                    \(L=64\) reaching equilibrium at \(T=2.36\) and
                \subref{sfig:equiandauto:autoM} the autocorrelation of an
                    Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                    sweeps -- otherwise the decline is too steep to show)
                    on half logarithmic axis.
                    The straight line is an exponential fit \(\exp(-t/\tau)\)
                    with \(\tau = 342(1)\).
            }
            \label{fig:equiandauto}
        \end{figure}\\
        Because every state is generated from the state before, measurements
        of subsequent states are correlated. To determine when two states
        are independent, one calculates the normalized autocorrelation function
        \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
        \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
        logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
        To get the autocorrelation time one can either fit a exponential
        function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
        or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
        \(\tau\) is an estimate after which time two samples are not
        correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
        To make sure that the error is not underestimated one should wait
        \(2\tau\) sweeps between two measurements.
        The autocorrelation time is of course dependent on the temperature.
        For example for the standard Metropolis algorithm the fluctuations
        are strong at high temperatures and subsequent
        states are more dissimilar and therefore less correlated than at low
        temperatures, where less spins are flipping. But the longest
        autocorrelation times are encountered at the critical temperature.
        This effect is called \emph{critical slowing down} and is
        characterized by the \emph{dynamical critical exponent} \(z\)
        \cite{SwendsenWang1987}. The dependence of the autocorrelation time
        \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).
        As mentioned in section \ref{sssec:wolff}, the Wolff cluster algorithm
        decreases \(z\) dramatically. This causes the course of the autocorrelation
        time in dependence on temperature to change significantly as shown in
        section \ref{sec:results}.
